Refunctionalization of Abstract Abstract Machines (Functional Pearl)

### 2018-07-05 paper/paper.tex
minor revisions

\subsubsection{Single-Step Transition}
Before describing how the machine evaluates expressions, we must first define several helper
functions. As mentioned in Section~\ref{anfsyntax}, atomic expressions are either a variable
or a literal @lambda@ term. As such, the atomic expression evaluator @atomicEval@ handles
these two cases and evaluates atomic expressions to closures in a straightforward way.
The @alloc@ function generates a fresh address, and always allocates a unique integer
in the domain of @store@.
The @isAtomic@ function is used as a predicate to determine if the expression is atomic.

@@

The above code shows the skeleton of @drive_step@.
At the end of the first case of pattern matching, we invoke @drive_step@ recursively 
on the new @NDState@.
\footnote{The code for constructing new environments, stores, and frames is elided for 
simplicity of presentation.}.
The code for the second case is replaced entirely by a function call to @continue@.
The shape of an interpreter in continuation-passing style has emerged, even though
the @continue@ function is not higher-order for the moment.

@@

We can now transform our abstract interpreter into cache-passing style. We begin by defining the 
return type @Ans@ as a case class which contains a @VS@ set and a cache. @Ans@ also implements 
its own accumulating operation @++@ which defines as a union of @Set[VS]@s and a join of caches.
The join operation of caches is defined as joining its two store components.

@@

Given a collection of elements of type @T@, an initial accumulator of type @Ans@,
a function @k@ that works on elements of type @T@, @nd@ chooses an element in
the collection and applies @k@ to it iteratively, and accumulates the value that
returned from @k@ when recuring on the rest of the collection. Eventually, @nd@
returns the accumulated value @acc@ when there is no element in the collection.
The latest accumulated cache @acc.cache@ is exposed to @k@ when applied.

@@

Now, as we can see, by using the @nd@ operator which abstracts over the meta-continuation,
the evaluation function can be simplified to one layer continuation-passing style.
The use two layers of continuations are unchained to two functions that each of them has one
continuation.
Now the continuation of @aeval@ plays the same role as the one in a concrete CPS interpreter.
Furthermore, we added a cache along with the abstract interpreter and @nd@ operator.

@@

The case where @e@ is a @let@ expression is not that obvious. As previously identified, 
function @nd@ works on a unknown number of nondeterministic choices, and it takes a 
function @k@ that works for every element in the collection. Here we present another
function @choices@ that uses delimited control operator @shift@ to capture that ``@k@''
when the program is direct-style and pass it to @nd@.

@@

For every nondeterministic choice in the abstract interpretation, we can simply call 
@choices@ on the set of possible choices and write the program sequentially as concrete 
interpreters, and the subsequent statements and expressions after the call-site become 
the delimited continuation @f@. We have two invocations of @choices@, which correspond 
to the two invocations of @nd@ in the refunctionalized form in the last section.
The function @choices@ also returns the latest cache to its left-hand side when invoked.
The cache will be still accumulated to the next call to @choices@ or @aeval@, 
given the fact that the abstract interpreter should always use the latest cache.

>>>>

\subsubsection{Single-Step Transition}
Before describing how the machine evaluates expressions, we must first define several helper
functions. As mentioned in Section~\ref{anfsyntax}, atomic expressions are either a variable
or a literal @lambda@ term. As such, the atomic expression evaluator @atomicEval@ handles
these two cases and evaluates atomic expressions to closures in a straightforward way.
The @alloc@ function generates a fresh address, and always allocates an unique integer
in the domain of @store@.
The @isAtomic@ function is used as a predicate to determine if the expression is atomic.

@@

The above code shows the skeleton of the @drive_step@.
At the end of the first case of pattern matching, we invoke @drive_step@ recursively 
on the new @NDState@.
\footnote{The code for constructing new environments, stores, and frames is elided for 
simplicity of presentation.}.
The code for the second case is replaced entirely by a function call to @continue@.
The shape of an interpreter in continuation-passing style has emerged, even though
the @continue@ function is not higher-order for the moment.

@@

We can now transform our abstract interpreter into cache-passing style. We begin by defining the 
return type @Ans@ as a case class which contains a @VS@ set and a cache. @Ans@ also implements 
its own accumulating operation @++@ which defines as an union of @Set[VS]@s and a join of caches.
The join operation of caches is defined as joining its two store components.

@@

Given a collection of elements of type @T@, an initial accumulator of type @Ans@,
a function @k@ that works on elements of type @T@, @nd@ chooses an element in
the collection and applies @k@ to it iteratively, and accumulates the value that
returned from @k@ when recurring on the rest of the collection. Eventually, @nd@
returns the accumulated value @acc@ when there is no element in the collection.
The latest accumulated cache @acc.cache@ is exposed to @k@ when applied.

@@

Now, as we can see, by using the @nd@ operator which abstracts over the meta-continuation,
the evaluation function can be simplified to one layer continuation-passing style.
The use of two layers of continuations are unchained to two functions that each of them has one
continuation.
The only continuation of @aeval@ plays the same role as the one in a concrete CPS interpreter.
Furthermore, we added a cache along with the abstract interpreter and the @nd@ operator.

@@

The case where @e@ is a @let@ expression is not that obvious. As previously identified, 
function @nd@ works on an unknown number of nondeterministic choices, and it takes a 
function @k@ that works for every element in the collection. Here we present another
function @choices@ that uses delimited control operator @shift@ to capture that ``@k@''
when the program is direct-style and pass it to @nd@.

@@

For every nondeterministic fork-point in the abstract interpretation, we can simply call 
@choices@ on the set of possible choices and write the program sequentially as concrete 
interpreters, and the subsequent statements and expressions after the call-site become 
the delimited continuation @f@. We have two invocations of @choices@, which correspond 
to the two invocations of @nd@ in the refunctionalized form in the last section.
The function @choices@ also returns the latest cache to its left-hand side when invoked.
The cache will be still accumulated to the next call to @choices@ or @aeval@, 
given the fact that the abstract interpreter should always use the latest cache.
